Does the COMPAS Needle Always Point Towards Equity? Finding Fairness in the COMPAS Risk Assessment Algorithm: A Case Study

### 2022-04-24 Paper_Draft2/Draft2.tex
fixing problems

% Affiliations / Addresses (Add [1] after \address if there is only one affiliation.)
\address{%
$^{1}$ \quad Statistical and Data Sciences Smith College Northampton, MA
01063; \href{mailto:bbaumer@smith.edu}{\nolinkurl{bbaumer@smith.edu}}\\
}
% Contact information of the corresponding author
\corres{Correspondence: \href{mailto:aacharya@smith.edu}{\nolinkurl{aacharya@smith.edu}},
\href{mailto:dcaravela@smith.edu}{\nolinkurl{dcaravela@smith.edu}},
\href{mailto:ekim89@smith.edu}{\nolinkurl{ekim89@smith.edu}},
\href{mailto:ekornberg@smith.edu}{\nolinkurl{ekornberg@smith.edu}},
\href{mailto:enesmith@smith.edu}{\nolinkurl{enesmith@smith.edu}}}

@@

The Correctional Offender Management Profiling for Alternative Sanctions
(COMPAS) algorithm was created by the private, for-profit company
Northpointe (now known by its parent company
\href{https://www.equivant.com/faq/}{equivant}), to predict defendants'
risk of recidivism. It generates a decile score that classifies
defendants' risk of recidivism as either low, medium, or high
\citep{angwin2016machine}. Jurisdictions across the United States use
the COMPAS risk assessment instrument, including but not limited to the
\href{https://doccs.ny.gov/system/files/documents/2020/11/8500.pdf}{New
York},
\href{https://hdsr.mitpress.mit.edu/pub/hzwo7ax4/release/4}{Massachusetts},
\href{https://hdsr.mitpress.mit.edu/pub/hzwo7ax4/release/4}{Michigan},
\href{https://hdsr.mitpress.mit.edu/pub/hzwo7ax4/release/4}{California},
and \href{https://doc.wi.gov/Pages/AboutDOC/COMPAS.aspx}{Wisconsin}
Departments of Corrections.

Due to the proprietary nature of the COMPAS algorithm, it is unknown how
exactly these recidivism risk scores are calculated; however, a
\href{https://www.documentcloud.org/documents/2702103-Sample-Risk-Assessment-COMPAS-CORE\#document/p5/a296598}{sample
COMPAS Risk Assessment Survey} has been made publicly available,
revealing the algorithm's input information. This survey has been widely
critiqued for using proxy variables for race that do not explicitly
factor in a defendant's race but heavily imply it, allowing Northpointe
to claim that their algorithm is free of racial bias. For example, the
COMPAS risk assessment survey asks screeners to speculate if a defendant
might be affiliated with a gang. It also asks if a defendant has any
friends or family members who have been crime victims
\citep{Angwin2016Sample}. Although these questions do not directly ask
about race, they do not take into account the pervasive nature of
systemic racism that infiltrates every aspect of the lives of
marginalized people, thereby indirectly asking about race.
\citet{angwin2016machine}. published a 2016 piece in the news
publication ProPublica that analyzes the methods and algorithms used by
Northpointe in their COMPAS risk score assessment algorithm and uncovers
the racial biases in defendants' scores \citep{angwin2016machine}. Their
research finds that ``the algorithm {[}is{]} somewhat more accurate than
a coin flip,'' a worrisome level of accuracy given the potential impact
its determinations may have on real people's lives.
\citet{angwin2016machine} specifically investigate the distribution of
COMPAS scores by decile among Black and white defendants. They write:
``The analysis also {[}shows{]} that even when controlling for prior
crimes, future recidivism, age, and gender, Black defendants {[}are{]}
45 percent more likely to be assigned higher risk scores than white
defendants'' \citep{larson2016we}. After examining the fairness metric
statistical parity difference, \citet{angwin2016machine} conclude that
the algorithm is racially biased \citep{larson2016we}.

@@

\href{https://www.womenatthetable.net/}{Women at the Table}, the sponsor
organization for this project, is ``a growing, global gender equality \&
democracy CSO based in Geneva, Switzerland focused on advancing feminist
systems change by using the prism of technology, innovation \& AI
exercising leverage points in technology, the economy, sustainability \&
democratic governance.'' We are collaborating with the organization on
its AI \& Equality \citep{noauthor_ai_nodate} initiative, tasked with
de-biasing the COMPAS algorithm \citep{aif360-oct-2018} and producing a
corresponding data story that will be added to its library.

Our project builds on Women at the Table's various de-biasing algorithms
used in its
\href{https://colab.research.google.com/drive/11SWijBDkr1pSgIXyjUkgoSB0gh5oAlaI}{AI
\& Equality Human Rights Toolbox} to conduct our own analyses on the
COMPAS data set. Based on this analysis, we employ a human rights
framework to contribute to the ProPublica and Northpointe debate and
investigate whether or to what extent there is racial bias in the COMPAS
algorithm. With a solid understanding of the two sides, we aim to
pinpoint the shortcomings of both arguments and correct them in our
analyses. We will use various de-biasing techniques and fairness metrics
to evaluate the level of bias present in the COMPAS data and our
algorithm. We will summarize our results using the JupyterNotebook
framework from Women at the Table, to be used by members of the
organization to teach in a workshop setting. We hope that our findings
will highlight the importance of checking statistical analyses using
varied methods and contribute to the ongoing discussion of the effects
of machine biases in the justice system.

@@

@@

@@

AI Fairness 360 (AIF360) is an open-source Python toolkit that seeks to
``to help facilitate the transition of fairness research algorithms to
use in an industrial setting and to provide a common framework for
fairness researchers to share and evaluate algorithms''
\citep{aif360-oct-2018}. It contains multiple data sets, including the
COMPAS data set that accompanied the \citet{angwin2016machine} piece.

>>>>

% Affiliations / Addresses (Add [1] after \address if there is only one affiliation.)
\address{%
$^{1}$ \quad Statistical and Data Sciences Smith College Northampton, MA 01063; \href{mailto:aacharya@smith.edu}{\nolinkurl{aacharya@smith.edu}},
\href{mailto:dcaravela@smith.edu}{\nolinkurl{dcaravela@smith.edu}},
\href{mailto:ekim89@smith.edu}{\nolinkurl{ekim89@smith.edu}},
\href{mailto:ekornberg@smith.edu}{\nolinkurl{ekornberg@smith.edu}},
\href{mailto:enesmith@smith.edu}{\nolinkurl{enesmith@smith.edu}}\\
}
% Contact information of the corresponding author
\corres{Correspondence: \href{mailto:bbaumer@smith.edu}{\nolinkurl{bbaumer@smith.edu}}}

@@

The Correctional Offender Management Profiling for Alternative Sanctions
(COMPAS) algorithm was created by the private, for-profit company
Northpointe (now known by its parent company
\href{https://www.equivant.com/faq/}{Equivant}), to predict defendants'
risk of recidivism. It generates a decile score that classifies
defendants' risk of recidivism as either low, medium, or high
\citep{angwin2016machine}. Jurisdictions across the United States use
the COMPAS risk assessment instrument, including but not limited to the
New York \citep{NYDOCCS2019}, Massachusetts, Michigan, California
\citep{Jackson2020Setting}, and Wisconsin \citep{WisconsinDOC}
Departments of Corrections.

Due to the proprietary nature of the COMPAS algorithm, it is unknown how
exactly these recidivism risk scores are calculated; however, a sample
COMPAS Risk Assessment Survey has been made publicly available,
revealing the algorithm's input information. This survey has been widely
critiqued for using proxy variables for race that do not explicitly
factor in a defendant's race but heavily imply it, allowing Northpointe
to claim that their algorithm is free of racial bias. For example, the
COMPAS risk assessment survey asks screeners to speculate if a defendant
might be affiliated with a gang. It also asks if a defendant has any
friends or family members who have been crime victims
\citep{Angwin2016Sample}. Although these questions do not directly ask
about race, they do not take into account the pervasive nature of
systemic racism that infiltrates every aspect of the lives of
marginalized people, thereby indirectly asking about race.
\citet{angwin2016machine}. published a 2016 piece in the news
publication ProPublica that analyzes the methods and algorithms used by
Northpointe in their COMPAS risk score assessment algorithm and uncovers
the racial biases in defendants' scores \citep{angwin2016machine}. Their
research finds that ``the algorithm {[}is{]} somewhat more accurate than
a coin flip,'' a worrisome level of accuracy given the potential impact
its determinations may have on real people's lives.
\citet{angwin2016machine} specifically investigate the distribution of
COMPAS scores by decile among Black and white defendants. They write:
``The analysis also {[}shows{]} that even when controlling for prior
crimes, future recidivism, age, and gender, Black defendants {[}are{]}
45 percent more likely to be assigned higher risk scores than white
defendants'' \citep{larson2016we}. After examining the fairness metric
statistical parity difference, \citet{angwin2016machine} conclude that
the algorithm is racially biased \citep{larson2016we}.

@@

\href{https://www.womenatthetable.net/}{Women at the Table}, the sponsor
organization for this project, is ``a growing, global gender equality \&
democracy CSO based in Geneva, Switzerland focused on advancing feminist
systems change by using the prism of technology, innovation \& AI
exercising leverage points in technology, the economy, sustainability \&
democratic governance.'' We are collaborating with the organization on
its AI \& Equality \citep{noauthor_ai_nodate} initiative, tasked with
de-biasing the COMPAS algorithm \citep{aif360-oct-2018} and producing a
corresponding data story that will be added to its library. Our project
builds on Women at the Table's various de-biasing algorithms used in its
AI \& Equality Human Rights Toolbox to conduct our own analyses on the
COMPAS data set. Based on this analysis, we employ a human rights
framework to contribute to the ProPublica and Northpointe debate and
investigate whether or to what extent there is racial bias in the COMPAS
algorithm. With a solid understanding of the two sides, we aim to
pinpoint the shortcomings of both arguments and correct them in our
analyses. We will use various de-biasing techniques and fairness metrics
to evaluate the level of bias present in the COMPAS data and our
algorithm. We will summarize our results using the JupyterNotebook
framework from Women at the Table, to be used by members of the
organization to teach in a workshop setting. We hope that our findings
will highlight the importance of checking statistical analyses using
varied methods and contribute to the ongoing discussion of the effects
of machine biases in the justice system.

@@

{\centering \includegraphics[width=1\linewidth]{../images/race_bar_plot_new}

\caption{This plot shows the distribution of defendant races. The dataset contains about 3200 Black defendants and 2100 white defendants. Therefore, there are more Black people represented in the dataset than white people.}\label{fig:race plot}
\end{figure}

@@

{\centering \includegraphics[width=1\linewidth]{../images/recid_bar_plot_new}

\caption{This plot shows the distribution of two year recidivism outcomes. Out of all defendants, there is a higher proportion of people who did not recidivate than who did. About 2500 people recidivated, whereas approximately 2700 did not.}\label{fig:recid plot}
\end{figure}

@@

AI Fairness 360 (AIF360) is an open-source Python toolkit that seeks to
``to help facilitate the transition of fairness research algorithms to
use in an industrial setting and to provide a common framework for
fairness researchers to share and evaluate algorithms''
\citep{aif360-oct-2018}. It contains multiple data sets, including the
COMPAS data set that accompanied the \citet{angwin2016machine} piece.
Using the \citet{reticulate} package, we conduct our analysis in an
RMarkdown document with Python code.
